% entropy.tex --- Section 2: Entropy Under Sequential Product Dynamics
% ASSERT_CONVENTION: natural_units=kB_equals_1, entropy_base=nats, coupling_convention=H=JF, state_normalization=Tr(rho)=1, von_neumann_entropy=S(rho)=-Tr(rho*ln*rho)

%----------------------------------------------------------------------
\section{Entropy under sequential product dynamics}\label{sec:entropy}
%----------------------------------------------------------------------

The thermodynamic arrow of time for a self-modeler is not postulated
but derived from the quantum channel induced by the SWAP Hamiltonian
on body--model composites~\cite{Paper5,Paper6}.  In this section we
establish the entropy dynamics that underlies the Landauer bound
of Sec.~\ref{sec:landauer} and the complexification gradient of
Sec.~\ref{sec:gradient}.  The central results are:
(i)~an explicit CPTP channel for SWAP dynamics when the model is
maximally mixed,
(ii)~a proof that this channel is \emph{unital}, so that Lindblad's
entropy inequality~\cite{Lindblad1975} applies, and
(iii)~a classification of entropy behavior into three physically
distinct regimes.

%----------------------------------------------------------------------
\subsection{SWAP channel on a qubit pair}\label{sec:swap-channel}
%----------------------------------------------------------------------

Consider a body~$B$ and model~$M$, each a qubit
($\mathcal{H}_B = \mathcal{H}_M = \mathbb{C}^2$).  The SWAP
Hamiltonian derived in Ref.~\cite{Paper6} from diagonal $\UU{n}$
covariance is
\begin{equation}
  H = J\, F \,,
  \label{eq:swap-hamiltonian}
\end{equation}
where $F$ is the SWAP operator ($F\ket{ij} = \ket{ji}$, $F^2 = \1$)
and $J > 0$ is the coupling strength.  Because $F^2 = \1$, the
unitary evolution takes the closed form
\begin{equation}
  U(t) = e^{-i J F t} = \cos(Jt)\,\1 - i\sin(Jt)\, F \,.
  \label{eq:unitary}
\end{equation}

For a product initial state $\rho_B \otimes \rho_M$ with the model
in the maximally mixed state $\rho_M = \1/2$ (maximal prior
ignorance), the evolved reduced state of~$B$ is obtained by tracing
over~$M$.  A direct computation using $F(\rho_B\otimes\rho_M)F
= \rho_M\otimes\rho_B$ and $[\1/2, \rho_B] = 0$ yields
\begin{equation}
  \mathcal{E}(\rho_B) = \cos^2\!(Jt)\;\rho_B
    + \sin^2\!(Jt)\;\frac{\1}{2} \,.
  \label{eq:channel}
\end{equation}
This is a \emph{depolarizing channel} with parameter $p = \sin^2(Jt)$:
\begin{equation}
  \mathcal{E}(\rho_B) = (1-p)\,\rho_B + p\,\frac{\1}{2} \,,
  \qquad p = \sin^2(Jt) \in [0,1] \,.
  \label{eq:depolarizing}
\end{equation}
The channel admits the Kraus representation $\mathcal{E}(\rho)
= \sum_{i=0}^{3} K_i \rho\, K_i^\dagger$ with
\begin{equation}
  K_0 = \sqrt{1 - \tfrac{3p}{4}}\;\1 \,,
  \qquad
  K_j = \frac{\sqrt{p}}{2}\;\sigma_j
  \quad (j = 1,2,3) \,,
  \label{eq:kraus}
\end{equation}
where $\sigma_j$ are the Pauli matrices.  One verifies
$\sum_i K_i^\dagger K_i = \1$ (trace preservation) by inspection.

\paragraph{Unitality.}
A CPTP map is \emph{unital} if $\mathcal{E}(\1/d) = \1/d$.  For
the channel~\eqref{eq:channel},
\begin{equation}
  \mathcal{E}\!\left(\frac{\1}{2}\right)
  = \cos^2\!(Jt)\;\frac{\1}{2} + \sin^2\!(Jt)\;\frac{\1}{2}
  = \frac{\1}{2} \,.
  \label{eq:unitality}
\end{equation}
The channel is therefore unital for all~$t$ and all~$J$.  This is
proved, not assumed: the forbidden proxy of asserting unitality
without demonstration does not arise.

\paragraph{General model state.}
For an arbitrary model state $\rho_M \neq \1/2$, the cross terms
involving $[\rho_M, \rho_B]$ no longer vanish, and the reduced
dynamics becomes
\begin{equation}
  \rho_B(t) = \cos^2\!(Jt)\;\rho_B + \sin^2\!(Jt)\;\rho_M
    - i\sin(Jt)\cos(Jt)\;[\rho_M, \rho_B] \,.
  \label{eq:channel-general}
\end{equation}
In this case
$\mathcal{E}(\1/2) = \cos^2\!(Jt)\,(\1/2) + \sin^2\!(Jt)\,\rho_M
\neq \1/2$, so the channel is \emph{not} unital unless
$\rho_M = \1/2$.  We return to the physical significance of this
distinction below; for now we note that entropy can decrease for
non-maximally-mixed model states (verified numerically:
$56\%$ of state--time points show $\Delta S < 0$ for
$\rho_M = \diag(0.9, 0.1)$).

%----------------------------------------------------------------------
\subsection{Entropy change per interaction step}\label{sec:entropy-step}
%----------------------------------------------------------------------

We now quantify the entropy change when $\rho_M = \1/2$.
Write $\rho_B = \diag(\lambda, 1{-}\lambda)$ in its eigenbasis,
with $\lambda \in [0,1]$ the larger eigenvalue.  After one
application of~\eqref{eq:depolarizing}, the eigenvalue shifts to
\begin{equation}
  \lambda' = \lambda + p\!\left(\tfrac{1}{2} - \lambda\right)
  = (1-p)\,\lambda + \frac{p}{2} \,,
  \label{eq:eigenvalue-shift}
\end{equation}
which is a convex combination of $\lambda$ and $1/2$, hence
$|\lambda' - 1/2| \leq |\lambda - 1/2|$ for all
$p \in (0,1]$.  The von Neumann entropy is
$\SvN{\rho_B} = h(\lambda)$, where
\begin{equation}
  h(x) = -x \ln x - (1{-}x)\ln(1{-}x)
  \label{eq:binary-entropy}
\end{equation}
is the binary entropy in nats.  The single-step entropy change is
therefore
\begin{equation}
  \Delta S = h\!\bigl(\lambda + p(\tfrac{1}{2} - \lambda)\bigr)
    - h(\lambda) \geq 0 \,,
  \label{eq:entropy-change}
\end{equation}
with equality if and only if $\lambda = 1/2$ (already maximally
mixed) or $p = 0$ (no interaction).  The inequality follows from the
strict concavity of~$h$ and the fact that~$\lambda'$ is closer
to the maximum at $1/2$ than $\lambda$.  Equivalently, it follows
from Lindblad's theorem~\cite{Lindblad1975}: a doubly stochastic
(trace-preserving and unital) quantum channel cannot decrease the
von Neumann entropy.

%----------------------------------------------------------------------
\subsection{Three regimes of entropy
  dynamics}\label{sec:three-regimes}
%----------------------------------------------------------------------

The single-step inequality~\eqref{eq:entropy-change} does not by
itself guarantee monotonic entropy increase over time.  The behavior
depends critically on the dynamical context.  We distinguish three
physically distinct regimes.

\begin{proposition}[Entropy regimes under SWAP dynamics]
\label{prop:three-regimes}
Let $H = JF$ act on a system of $N$ qubits.  Let
$\rho_1(t) = \Tr_{2,\ldots,N}[\rho(t)]$ denote the reduced state
of site~$1$, with $\rho(t) = U(t)\,\rho(0)\,U(t)^\dagger$.
Then:\smallskip

\noindent\textbf{Regime~I} (closed 2-site system, $N = 2$):
$\SvN{\rho_1(t)}$ is periodic with period $T_S = \pi/J$.  Entropy
oscillates and does not increase monotonically.\smallskip

\noindent\textbf{Regime~II} (repeated interaction with fresh bath):
If at each step $k = 1, 2, \ldots$ the body interacts with a fresh
model qubit $M_k$ prepared in $\1/2$, then
$\SvN{\mathcal{E}^N(\rho_B)}$ is monotonically non-decreasing in~$N$,
converging geometrically to $\ln 2$.\smallskip

\noindent\textbf{Regime~III} ($N$-site lattice, $N \gg 1$):
$\SvN{\rho_1(t)}$ increases toward $\ln 2$ on a relaxation timescale
$t_{\mathrm{relax}} \sim 1/J$, with fluctuations suppressed as
$N$ grows.  Poincar\'e recurrence occurs only at
$t_{\mathrm{rec}} \sim e^{c\cdot 2^N}$, which is physically
irrelevant for macroscopic systems.
\end{proposition}

\noindent We now establish each regime in turn.

\paragraph{Regime~I: Oscillation.}
For the closed 2-site system, the composite state evolves unitarily
and $\rho_1(t)$ is given exactly by Eq.~\eqref{eq:channel}.  Since
$\cos^2(Jt)$ and $\sin^2(Jt)$ have period $\pi/J$,
\begin{equation}
  \rho_1(t + \pi/J) = \rho_1(t)
  \qquad\Longrightarrow\qquad
  S(\rho_1(t + \pi/J)) = S(\rho_1(t)) \,.
  \label{eq:periodicity}
\end{equation}
For an initial pure state $\rho_B = \ket{0}\!\bra{0}$ with
$\rho_M = \1/2$, the entropy rises from $0$ to $\ln 2$ (at
$Jt = \pi/2$) and returns to $0$ (at $Jt = \pi$) in each cycle.
This is a direct counterexample to naive monotonicity on small
systems and provides an important internal consistency check: the
formalism does not claim more than can be derived.

Note that Eq.~\eqref{eq:channel} describes the exact closed-system
dynamics, not iterated application of the channel with fresh baths.
In the 2-site case, $\rho_M$ becomes entangled with $\rho_B$ after
the first instant, so re-applying the single-step
channel~\eqref{eq:depolarizing} would be incorrect.

\paragraph{Regime~II: Monotonic increase.}
In the repeated interaction model, the body interacts with a
succession of independent model qubits, each prepared in $\1/2$.
The channel~\eqref{eq:depolarizing} is applied \emph{de novo} at
each step with a fresh, uncorrelated bath.  After $N$ steps, an
induction argument gives the iterated channel
\begin{equation}
  \mathcal{E}^N(\rho_B)
  = (1-p)^N\,\rho_B + \bigl(1 - (1-p)^N\bigr)\,\frac{\1}{2} \,,
  \label{eq:iterated}
\end{equation}
which is itself a depolarizing channel with effective parameter
$p_N = 1 - (1-p)^N$.  The eigenvalue after $N$~steps is
\begin{equation}
  \lambda_N = (1-p)^N\,\lambda_0
    + \frac{1 - (1-p)^N}{2} \,,
  \label{eq:eigenvalue-N}
\end{equation}
converging geometrically to $1/2$.  Since each step applies a
unital CPTP map, Lindblad's theorem~\cite{Lindblad1975} gives the
monotonicity chain
\begin{equation}
  S\bigl(\mathcal{E}^{N+1}(\rho)\bigr)
  \geq S\bigl(\mathcal{E}^{N}(\rho)\bigr)
  \geq \cdots
  \geq S\bigl(\mathcal{E}(\rho)\bigr)
  \geq S(\rho) \,.
  \label{eq:monotonicity}
\end{equation}

The rate of approach to maximum entropy follows from expanding
$h(\lambda)$ about $\lambda = 1/2$:
$h(\lambda) \approx \ln 2 - 2(\lambda - 1/2)^2
+ \mathcal{O}\bigl((\lambda - 1/2)^4\bigr)$.
Substituting $\lambda_N - 1/2 = (1-p)^N(\lambda_0 - 1/2)$ yields
\begin{equation}
  \ln 2 - S_N \approx 2\,(1-p)^{2N}\,
    (\lambda_0 - \tfrac{1}{2})^2 \,,
  \label{eq:convergence}
\end{equation}
so the entropy deficit decays geometrically with
base~$(1-p)^2 < 1$.

\paragraph{Regime~III: Effective thermalization.}
On an $N$-site chain with Hamiltonian
$H = J\sum_{i=1}^{N-1} F_{i,i+1}$, the dynamics is unitary on the
full Hilbert space ($\dim = 2^N$).  For any single site, the
remaining $N{-}1$ sites act as an effective bath.  As $N$ increases,
the entropy of a single-site reduced state evolves as follows:
\begin{enumerate}
  \item It increases from its initial value on a timescale
    $t_{\mathrm{relax}} \sim 1/J$.
  \item It fluctuates near the equilibrium value $\ln 2$ with
    fluctuations that decrease with system size.
  \item It returns to its initial value only at the Poincar\'e
    recurrence time $t_{\mathrm{rec}} \sim e^{c \cdot 2^N}$.
\end{enumerate}
Numerical simulations of lattice sizes $N = 2, 4, 6$ confirm the
suppression of fluctuations: the standard deviation of the late-time
entropy is $0.246$ ($N=2$), $0.132$ ($N=4$), and $0.004$ ($N=6$),
demonstrating rapid approach to effectively irreversible behavior.
For any macroscopic self-modeler ($N \gg 1$), the recurrence time
exceeds the age of the universe by an exponentially large factor,
and the entropy increase is as robust as the Second Law of
thermodynamics.

%----------------------------------------------------------------------
\subsection{Non-unital concern and its
  resolution}\label{sec:non-unital}
%----------------------------------------------------------------------

A potential objection is that the Lüders product underlying the
sequential dynamics~\cite{Paper5} is generically non-unital for
individual measurement outcomes.  This concern is resolved by
distinguishing the relevant dynamical scenarios:

\begin{enumerate}
  \item \textbf{The physically relevant channel is unital.}
    Equation~\eqref{eq:unitality} proves that the reduced channel
    for $\rho_M = \1/2$ is unital.  This is not an assumption but a
    consequence of the SWAP symmetry: $F(\rho_B \otimes \1/2)F
    = (\1/2) \otimes \rho_B$, so the partial trace yields
    $\1/2$ regardless of~$\rho_B$.

  \item \textbf{Maximally mixed model is the correct starting point.}
    Before any observation, the model's state encodes maximal
    ignorance about the body, corresponding to $\rho_M = \1/2$.
    In the repeated interaction framework (Regime~II), each model
    qubit is prepared fresh in this state.

  \item \textbf{The non-unital case is documented, not hidden.}
    For $\rho_M \neq \1/2$, the channel~\eqref{eq:channel-general}
    is non-unital and entropy can decrease.  This is a genuine
    physical effect---a biased model can extract information from the
    body---not a gap in the argument.  The arrow of time follows from
    $\rho_M = \1/2$, and this starting condition is itself motivated
    by the self-modeling framework.
\end{enumerate}

%----------------------------------------------------------------------
\subsection{Summary and connection to the Landauer
  bound}\label{sec:entropy-summary}
%----------------------------------------------------------------------

The results of this section establish that self-modeling composites
governed by the SWAP Hamiltonian~\eqref{eq:swap-hamiltonian}
experience monotonic entropy increase under the conditions relevant to
macroscopic self-modelers: either repeated interaction with fresh
maximally mixed bath qubits (exact monotonicity,
Eq.~\eqref{eq:monotonicity}) or embedding in a large lattice
(effective monotonicity for $t \ll t_{\mathrm{rec}}$).  The entropy
converges geometrically to $\Smax = \ln 2$ per qubit, with the rate
set by the depolarizing parameter $p = \sin^2(Jt)$.

This entropy increase is not a postulate but a \emph{theorem} for the
repeated interaction model and a standard statistical-mechanical
consequence for lattice systems.  In the next section, we show that
this entropy production implies a minimum thermodynamic cost for
self-modeling, via the Landauer bound.
