% ASSERT_CONVENTION: entropy_base=nats, mutual_information=H(B)+H(M)-H(B,M), experiential_density=I*(1-I/H(B)), matrix_norm=row_sum_inf, l1_not_tv=true
%
% Lipschitz Stability of the Experiential Density
% Phase 02, Plan 01: Formal proof with explicit constants
%
% Conventions:
%   - All entropies in nats (natural logarithm)
%   - ||.||_1 = sum |p_i - q_i| throughout (NOT total variation)
%   - ||P||_inf = max_i sum_j |P_ij| (row-sum norm)
%   - Stationary distribution: row vector pi with pi P = pi

\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{definition}[theorem]{Definition}

\DeclareMathOperator{\gap}{gap}
\DeclareMathOperator{\tr}{Tr}

\title{Lipschitz Stability of the Experiential Density}
\author{Phase 02, Plan 01}
\date{}

\begin{document}
\maketitle

%% ====================================================================
%% Section 1: Setup and Statement
%% ====================================================================
\section{Setup and Theorem Statement}

\begin{definition}[Experiential density]
Let $P$ be an irreducible row-stochastic matrix on $\Omega = B \times M$
with stationary distribution $\pi$ (the unique row vector satisfying
$\pi P = \pi$, $\pi_i \ge 0$, $\sum_i \pi_i = 1$).
Reshape $\pi$ as a joint distribution $p(b,m)$ on $B \times M$.
The \emph{experiential density} is
\begin{equation}\label{eq:rho-def}
  \rho(P) \;=\; I_\pi(B;M)\,\Bigl(1 - \frac{I_\pi(B;M)}{H_\pi(B)}\Bigr),
\end{equation}
where $I_\pi(B;M) = H_\pi(B) + H_\pi(M) - H_\pi(B,M)$ is the mutual
information and $H_\pi(B) = -\sum_b p_B(b)\ln p_B(b)$ is the marginal
entropy on $B$, all in nats.
\end{definition}

This functional belongs to the LMC (Lopez-Ruiz--Mancini--Calbet) statistical
complexity family~\cite{LMC1995}, which measures the product of information
content and disequilibrium.  More broadly, it sits within the effective
complexity framework of Gell-Mann and Lloyd~\cite{GellMannLloyd2003},
capturing the idea that meaningful complexity peaks between pure order
($I = H(B)$, $\rho = 0$) and pure randomness ($I = 0$, $\rho = 0$).

\begin{definition}[Spectral gap and norms]
For an irreducible stochastic matrix $P$ on $n = |\Omega|$ states:
\begin{itemize}
  \item The \emph{absolute spectral gap} is
        $\gap(P) = 1 - |\lambda_2(P)|$, where $\lambda_2$ is the
        eigenvalue of $P$ with second-largest absolute value.
  \item The \emph{row-sum norm} is
        $\|P\|_\infty = \max_i \sum_j |P_{ij}|$.
        For the difference of stochastic matrices,
        $\|P - P'\|_\infty = \max_i \sum_j |P_{ij} - P'_{ij}|
        = \max_i \|P_i - P'_i\|_1$.
  \item We use the $\ell^1$ norm throughout:
        $\|p - q\|_1 = \sum_i |p_i - q_i|$.
        The total variation distance is
        $\|p-q\|_{\mathrm{TV}} = \tfrac{1}{2}\|p-q\|_1$.
\end{itemize}
\end{definition}

\begin{theorem}[Lipschitz stability of $\rho$]\label{thm:lipschitz}
Let $P$, $P'$ be irreducible row-stochastic matrices on
$\Omega = B \times M$ with $|B| \ge 2$, $|M| \ge 2$, $|\Omega| = n$.
Let $\pi$, $\pi'$ be their stationary distributions, and suppose
$\gap(P) > 0$.

\medskip\noindent
\textbf{Non-linear bound (sharpest form).}
Define $\delta = \|\pi - \pi'\|_1$.  Then:
\begin{equation}\label{eq:rho-nonlinear}
  |\rho(P) - \rho(P')| \;\le\;
  \frac{\delta}{2}\bigl[2\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
  + 4\,h_{\mathrm{bin}}\!\Bigl(\frac{\delta}{2}\Bigr),
\end{equation}
where $h_{\mathrm{bin}}(x) = -x\ln x - (1-x)\ln(1-x)$ and
$\delta \le \|P - P'\|_\infty / \gap(P)$ by the Cho--Meyer bound.

\medskip\noindent
\textbf{Linear Lipschitz bound.}
Equivalently,
\begin{equation}\label{eq:lipschitz-bound}
  |\rho(P) - \rho(P')| \;\le\; L(\delta) \cdot \|P - P'\|_\infty,
\end{equation}
where
\begin{equation}\label{eq:L-explicit}
  L(\delta) \;=\; \frac{C_I(\delta) + C_H(\delta)}{\gap(P)},
\end{equation}
with
\begin{align}
  C_I(\delta) &= \frac{1}{2}\bigl[\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
         + 3\,\frac{h_{\mathrm{bin}}(\delta/2)}{\delta}, \label{eq:CI} \\
  C_H(\delta) &= \frac{1}{2}\ln(|B|-1) + \frac{h_{\mathrm{bin}}(\delta/2)}{\delta}. \label{eq:CH}
\end{align}

For small perturbations ($\delta \ll 1$),
$h_{\mathrm{bin}}(\delta/2)/\delta \sim \frac{1}{2}\ln(2/\delta)$,
which is large but finite for any fixed $\delta > 0$.
The leading behavior of $L$ is:
\begin{equation}\label{eq:L-leading}
  L \;\sim\; \frac{1}{2\,\gap(P)}\,
  \bigl[2\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
  + \frac{2\ln(2/\delta)}{\gap(P)}.
\end{equation}

\medskip\noindent
\textbf{Uniform non-linear bound (for all perturbation sizes simultaneously).}
Substituting $\delta \le \|P-P'\|_\infty/\gap(P)$ into~\eqref{eq:rho-nonlinear}
and using $h_{\mathrm{bin}}(x) \le \ln 2$ for all $x$:
\begin{equation}\label{eq:rho-uniform}
  |\rho(P) - \rho(P')| \;\le\;
  \frac{\|P-P'\|_\infty}{2\,\gap(P)}\bigl[2\ln(|B|-1) + \ln(|M|-1)
  + \ln(n-1)\bigr] + 4\ln 2.
\end{equation}
The first term is the Lipschitz part (linear in $\|P-P'\|_\infty$)
and the second is a universal additive constant ($4\ln 2 \approx 2.77$ nats).
For large perturbations, the Lipschitz term dominates; for small
perturbations, the additive constant is an overestimate (the true
bound~\eqref{eq:rho-nonlinear} is much smaller).

\medskip\noindent
\textbf{Domain of validity:}
\begin{enumerate}
  \item $P$ irreducible with $\gap(P) > 0$;
  \item $P'$ irreducible (so $\pi'$ exists and is unique);
  \item $|B| \ge 2$ (otherwise $H_\pi(B) = 0$ and $\rho \equiv 0$).
\end{enumerate}
No lower bound on $H_\pi(B)$ is required in the Lipschitz constant.
\end{theorem}

The proof proceeds in three steps, each using a standard result.

%% ====================================================================
%% Section 2: Step 1 -- Stationary Distribution Perturbation
%% ====================================================================
\section{Step 1: Stationary Distribution Perturbation}

\begin{lemma}[Cho--Meyer perturbation bound~\cite{ChoMeyer2001}]
\label{lem:cho-meyer}
Let $P$, $P'$ be irreducible stochastic matrices on $n$ states with
stationary distributions $\pi$, $\pi'$.  Let
$\Pi = \mathbf{1}\pi$ be the rank-one stationary projector.  Then
\begin{equation}\label{eq:cho-meyer}
  \|\pi - \pi'\|_1 \;\le\;
  \frac{1}{\gap(P)}\,\|P - P'\|_\infty.
\end{equation}
\end{lemma}

\begin{proof}[Proof sketch (following Cho--Meyer~\cite{ChoMeyer2001})]
From $\pi P = \pi$ and $\pi' P' = \pi'$:
\[
  \pi' - \pi \;=\; \pi'(P' - P)(I - P + \Pi)^{-1},
\]
where $(I - P + \Pi)^{-1}$ is the group inverse of $(I - P)$.
The key bound is $\|(I - P + \Pi)^{-1}\|_\infty \le 1/\gap(P)$
(Meyer~\cite{Meyer1975}; Cho--Meyer~\cite{ChoMeyer2001}, Theorem~3.1).
Since $\|\pi'\|_1 = 1$:
\[
  \|\pi' - \pi\|_1
  \;\le\; \|\pi'\|_1 \cdot \|P' - P\|_\infty \cdot \|(I - P + \Pi)^{-1}\|_\infty
  \;\le\; \frac{\|P - P'\|_\infty}{\gap(P)}.
\]
Here we used $\|v A\|_1 \le \|v\|_1 \cdot \|A\|_\infty$ for a row
vector $v$ and matrix $A$ (standard submultiplicativity for the
$\ell^1$-$\ell^\infty$ pair).
\end{proof}

\begin{remark}
The bound~\eqref{eq:cho-meyer} may be loose for specific chains.
The sharper Cho--Meyer condition number
$\kappa(P) = \|\pi\|_\infty \cdot \|(I - P + \Pi)^{-1}\|_\infty$
can be much smaller than $1/\gap(P)$.  We use the simpler bound because
it gives clean dependence on the spectral gap.
\end{remark}

\textbf{Interface to Step 2:}
Define $\delta = \|\pi - \pi'\|_1$.  Reshaping $\pi$ and $\pi'$ as
joint distributions $p$ and $q$ on $B \times M$ (with
$p(b,m) = \pi_{(b,m)}$), we have $\|p - q\|_1 = \|\pi - \pi'\|_1 = \delta$.
By Lemma~\ref{lem:cho-meyer}:
\begin{equation}\label{eq:delta-bound}
  \delta \;\le\; \frac{\|P - P'\|_\infty}{\gap(P)}.
\end{equation}

%% ====================================================================
%% Section 3: Step 2 -- Mutual Information Continuity
%% ====================================================================
\section{Step 2: Mutual Information and Entropy Continuity}

\subsection{Entropy continuity lemma}

\begin{lemma}[Entropy continuity~\cite{CsiszarKorner2011,Audenaert2007}]
\label{lem:entropy-cont}
Let $p$, $q$ be probability distributions on an alphabet of size $k \ge 2$
with $\|p - q\|_1 = \delta \in (0, 2(1 - 1/k)]$.  Then
\begin{equation}\label{eq:fannes}
  |H(p) - H(q)| \;\le\; \frac{\delta}{2}\,\ln(k-1)
  + h_{\mathrm{bin}}\!\Bigl(\frac{\delta}{2}\Bigr),
\end{equation}
where $h_{\mathrm{bin}}(x) = -x\ln x - (1-x)\ln(1-x)$ in nats.
\end{lemma}

\begin{remark}[Convention check]\label{rem:convention}
Many references state this bound with $\delta_{\mathrm{TV}} = \|p-q\|_{\mathrm{TV}}
= \frac{1}{2}\|p-q\|_1$.  In that convention, the bound reads
$|H(p) - H(q)| \le \delta_{\mathrm{TV}}\ln(k-1) + h_{\mathrm{bin}}(\delta_{\mathrm{TV}})$.
Our formula~\eqref{eq:fannes} is the same after substituting
$\delta_{\mathrm{TV}} = \delta/2$.
The tight constant is due to Audenaert~\cite{Audenaert2007}.
\end{remark}

\subsection{Data processing inequality for marginals}

\begin{lemma}[Marginal contraction]\label{lem:marginal}
Let $p$, $q$ be joint distributions on $B \times M$.  Then
\begin{equation}\label{eq:marginal-contract}
  \|p_B - q_B\|_1 \;\le\; \|p - q\|_1,
  \qquad
  \|p_M - q_M\|_1 \;\le\; \|p - q\|_1,
\end{equation}
where $p_B(b) = \sum_m p(b,m)$ and $p_M(m) = \sum_b p(b,m)$.
\end{lemma}

\begin{proof}
By the triangle inequality:
$\|p_B - q_B\|_1 = \sum_b |p_B(b) - q_B(b)|
= \sum_b \bigl|\sum_m (p(b,m) - q(b,m))\bigr|
\le \sum_b \sum_m |p(b,m) - q(b,m)| = \|p - q\|_1$.
\end{proof}

\subsection{Mutual information continuity bound}

Using $I(B;M) = H(B) + H(M) - H(B,M)$ and the triangle inequality:
\begin{equation}\label{eq:MI-triangle}
  |I_p(B;M) - I_q(B;M)| \;\le\;
  |H_p(B) - H_q(B)| + |H_p(M) - H_q(M)| + |H_p(B,M) - H_q(B,M)|.
\end{equation}

Apply Lemma~\ref{lem:entropy-cont} to each term.  The marginal
perturbation bounds (Lemma~\ref{lem:marginal}) give
$\|p_B - q_B\|_1 \le \delta$ and $\|p_M - q_M\|_1 \le \delta$,
where $\delta = \|p - q\|_1$.

\begin{itemize}
  \item $H(B)$: alphabet size $|B|$, perturbation $\le \delta$:
    $|H_p(B) - H_q(B)| \le \frac{\delta}{2}\ln(|B|-1) + h_{\mathrm{bin}}(\delta/2)$.

  \item $H(M)$: alphabet size $|M|$, perturbation $\le \delta$:
    $|H_p(M) - H_q(M)| \le \frac{\delta}{2}\ln(|M|-1) + h_{\mathrm{bin}}(\delta/2)$.

  \item $H(B,M)$: alphabet size $n = |B| \cdot |M|$, perturbation $= \delta$:
    $|H_p(B,M) - H_q(B,M)| \le \frac{\delta}{2}\ln(n-1) + h_{\mathrm{bin}}(\delta/2)$.
\end{itemize}

Summing:
\begin{equation}\label{eq:MI-bound}
  |I_p(B;M) - I_q(B;M)| \;\le\;
  \frac{\delta}{2}\bigl[\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
  + 3\,h_{\mathrm{bin}}\!\Bigl(\frac{\delta}{2}\Bigr).
\end{equation}

\subsection{Entropy of $B$ continuity (separate bound for Step 3)}

We also need $|H_p(B) - H_q(B)|$ separately:
\begin{equation}\label{eq:HB-bound}
  |H_p(B) - H_q(B)| \;\le\;
  \frac{\delta}{2}\ln(|B|-1) + h_{\mathrm{bin}}\!\Bigl(\frac{\delta}{2}\Bigr).
\end{equation}

%% ====================================================================
%% Section 4: Step 3 -- Composition through rho = I(1 - I/H)
%% ====================================================================
\section{Step 3: Composition through $\rho = I(1 - I/H)$}

Write $\rho = f(I, H)$ where $f(I, H) = I(1 - I/H)$, with
$I = I_\pi(B;M) \ge 0$, $H = H_\pi(B) > 0$, and $0 \le I \le H$.

\subsection{Gradient bound via mean value theorem}

The partial derivatives of $f$ on the domain $\mathcal{D} = \{(I,H) : 0 \le I \le H,\, H > 0\}$:
\begin{align}
  \frac{\partial f}{\partial I} &= 1 - \frac{2I}{H},
  \qquad\text{so } \Bigl|\frac{\partial f}{\partial I}\Bigr| \le 1
  \quad\text{(since $I/H \in [0,1]$)}; \label{eq:df-dI}\\[4pt]
  \frac{\partial f}{\partial H} &= \frac{I^2}{H^2},
  \qquad\text{so } \Bigl|\frac{\partial f}{\partial H}\Bigr| \le 1
  \quad\text{(since $I/H \in [0,1]$)}. \label{eq:df-dH}
\end{align}

The domain $\mathcal{D}$ is convex: for $(I,H), (I',H') \in \mathcal{D}$
and $t \in [0,1]$, the convex combination
$\bar{H} = (1-t)H + tH' > 0$ and
$\bar{I} = (1-t)I + tI' \le (1-t)H + tH' = \bar{H}$, so
$(\bar{I},\bar{H}) \in \mathcal{D}$.

By the multivariate mean value theorem, there exists
$(\bar{I},\bar{H})$ on the line segment from $(I,H)$ to $(I',H')$
such that:
\begin{align}
  f(I,H) - f(I',H') &= \frac{\partial f}{\partial I}\Big|_{(\bar{I},\bar{H})}
  (I - I') + \frac{\partial f}{\partial H}\Big|_{(\bar{I},\bar{H})}(H - H'). \notag
\end{align}

Therefore:
\begin{equation}\label{eq:mvt}
  |f(I,H) - f(I',H')|
  \;\le\; \Bigl|\frac{\partial f}{\partial I}\Bigr|\,|I - I'|
  + \Bigl|\frac{\partial f}{\partial H}\Bigr|\,|H - H'|
  \;\le\; |I - I'| + |H - H'|.
\end{equation}

\begin{remark}
This bound is clean: no factors of $H$, $H'$, or $h_{\min}$ appear.
The key observation is that $f(I,H) = I(1-I/H)$ has
$\|\nabla f\|_\infty \le 1$ everywhere on its feasible domain,
because $I/H$ is bounded in $[0,1]$.
\end{remark}

\subsection{Final composition}

Combining~\eqref{eq:mvt} with the MI bound~\eqref{eq:MI-bound},
the $H(B)$ bound~\eqref{eq:HB-bound}, and the Cho--Meyer
bound~\eqref{eq:delta-bound}:

\begin{align}
  |\rho(P) - \rho(P')|
  &\le |I - I'| + |H - H'| \notag\\
  &\le \biggl[\frac{\delta}{2}\bigl(\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr)
       + 3\,h_{\mathrm{bin}}\!\Bigl(\frac{\delta}{2}\Bigr)\biggr] \notag\\
  &\quad + \biggl[\frac{\delta}{2}\ln(|B|-1)
       + h_{\mathrm{bin}}\!\Bigl(\frac{\delta}{2}\Bigr)\biggr] \notag\\
  &= \frac{\delta}{2}\bigl[2\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
     + 4\,h_{\mathrm{bin}}\!\Bigl(\frac{\delta}{2}\Bigr), \label{eq:rho-bound-nonlinear}
\end{align}
with $\delta \le \|P - P'\|_\infty / \gap(P)$.

This is the non-linear bound~\eqref{eq:rho-nonlinear}.

\medskip
\textbf{Linear Lipschitz form.}  Writing
$h_{\mathrm{bin}}(\delta/2) \le \delta \cdot [h_{\mathrm{bin}}(\delta/2)/\delta]$
and defining $\eta(\delta) = h_{\mathrm{bin}}(\delta/2)/\delta$ (finite for
any $\delta > 0$):
\begin{align}
  |\rho(P) - \rho(P')|
  &\le \bigl[C_I(\delta) + C_H(\delta)\bigr]\,\delta \notag\\
  &\le \frac{C_I(\delta) + C_H(\delta)}{\gap(P)}\,\|P - P'\|_\infty,
  \label{eq:rho-bound-linear}
\end{align}
where $C_I(\delta)$ and $C_H(\delta)$ are as in~\eqref{eq:CI}--\eqref{eq:CH}
with $\eta = \eta(\delta) = h_{\mathrm{bin}}(\delta/2)/\delta$.

\begin{equation}\label{eq:L-final}
  \boxed{L(\delta) \;=\; \frac{C_I(\delta) + C_H(\delta)}{\gap(P)}}
\end{equation}

\begin{remark}[On the $\delta$-dependence of $L$]
The Lipschitz ``constant'' $L(\delta)$ depends on $\delta$ through
the $h_{\mathrm{bin}}/\delta$ ratio, which grows as $\frac{1}{2}\ln(2/\delta)$
for small $\delta$.  This is not a deficiency of the proof: it reflects
a genuine feature of entropy continuity.  Shannon entropy is \emph{not}
Lipschitz (in the usual sense of a finite global constant) on the
probability simplex, because $|H(p) - H(q)|/\|p-q\|_1$ diverges
logarithmically as $p$ approaches a vertex of the simplex.

For any \emph{fixed} perturbation size $\varepsilon = \|P - P'\|_\infty$,
the constant $L(\varepsilon/\gap(P))$ is finite and explicit.
For the non-linear bound~\eqref{eq:rho-nonlinear}, no $\delta$-dependent
constant is needed: the bound is valid for all $\delta$ simultaneously.
\end{remark}

\begin{remark}[Improvement over the plan's estimate]
The plan estimated $L = (2C_I + C_H)/\gap(P)$ using a product-rule
decomposition of $|f(I,H) - f(I',H')|$.  The mean value theorem
yields the tighter bound $L = (C_I + C_H)/\gap(P)$, replacing
the factor $2C_I$ with $C_I$.  This is because $|\partial f/\partial I| \le 1$
(not $\le 2$) on the feasible domain.
\end{remark}

%% ====================================================================
%% Section 5: Explicit Formulas
%% ====================================================================
\section{Explicit Formulas}

\textbf{Notation:} $n = |B| \cdot |M| = |\Omega|$,
$\gap = \gap(P)$, $\varepsilon = \|P - P'\|_\infty$,
$\delta = \|\pi - \pi'\|_1 \le \varepsilon/\gap$.

\medskip\noindent
\textbf{Non-linear bound (recommended for evaluation):}
\begin{equation}\label{eq:nonlinear-final}
  |\rho(P) - \rho(P')| \;\le\;
  \frac{\varepsilon}{2\,\gap}\bigl[2\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
  + 4\,h_{\mathrm{bin}}\!\Bigl(\frac{\varepsilon}{2\,\gap}\Bigr).
\end{equation}

\noindent
\textbf{Linear bound (with $\delta$-dependent constant):}
\begin{align}
  L(\delta) &= \frac{1}{\gap}\left\{
    \frac{1}{2}\bigl[2\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
    + 4\,\frac{h_{\mathrm{bin}}(\delta/2)}{\delta}
  \right\}. \label{eq:L-collected}
\end{align}

\noindent
\textbf{Asymptotic form (small perturbation, $\delta \to 0$):}
\begin{equation}\label{eq:L-asymptotic}
  L \;\sim\; \frac{1}{2\,\gap}\bigl[2\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
  + \frac{2}{\gap}\ln\!\Bigl(\frac{2\,\gap}{\varepsilon}\Bigr).
\end{equation}

\noindent
\textbf{Uniform additive bound (for all perturbation sizes):}
\begin{equation}\label{eq:uniform}
  |\rho(P) - \rho(P')| \;\le\;
  \frac{\varepsilon}{2\,\gap}\bigl[2\ln(|B|-1) + \ln(|M|-1) + \ln(n-1)\bigr]
  + 4\ln 2,
\end{equation}
using $h_{\mathrm{bin}}(x) \le \ln 2$ for all $x \in [0,1]$.

%% ====================================================================
%% Bibliography
%% ====================================================================

\begin{thebibliography}{99}
\bibitem{LMC1995}
R.~Lopez-Ruiz, H.~L.~Mancini, X.~Calbet,
``A statistical measure of complexity,''
\emph{Phys.\ Lett.\ A} \textbf{209} (1995) 321--326.

\bibitem{GellMannLloyd2003}
M.~Gell-Mann, S.~Lloyd,
``Effective complexity,''
in \emph{Nonextensive Entropy}, eds.\ M.~Gell-Mann, C.~Tsallis,
Oxford Univ.\ Press (2004); original preprint 1996/2003.

\bibitem{ChoMeyer2001}
G.~E.~Cho, C.~D.~Meyer,
``Comparison of perturbation bounds for the stationary distribution
of a Markov chain,''
\emph{Linear Algebra Appl.} \textbf{335} (2001) 137--150.

\bibitem{Meyer1975}
C.~D.~Meyer,
``The role of the group generalized inverse in the theory of
finite Markov chains,''
\emph{SIAM Rev.} \textbf{17} (1975) 443--464.

\bibitem{CsiszarKorner2011}
I.~Csisz\'ar, J.~K\"orner,
\emph{Information Theory: Coding Theorems for Discrete Memoryless Systems},
2nd ed., Cambridge Univ.\ Press (2011), Lemma~2.7.

\bibitem{Audenaert2007}
K.~M.~R.~Audenaert,
``A sharp continuity estimate for the von Neumann entropy,''
\emph{J.\ Phys.\ A} \textbf{40} (2007) 8127--8136;
arXiv:quant-ph/0610146.
\end{thebibliography}

\end{document}
